---
title: "Florida Census Demographics (2000–2020)"
---

```{r}
#| label: setup
#| message: false
#| echo: false

library(tidyverse)
library(sf)
library(leaflet)

# Load county boundaries and historical decennial census counts
fl_geo <- read_rds("data/fl_geo.rds")
fl_race <- read_rds("data/fl_race_2000_2020.rds")

# Calculate Hispanic population percentage and 20-year shift
change_hisp <- fl_race |>
  filter(variable == "hispanic") |>
  mutate(pct = round(100 * (value / summary_value), 1)) |>
  select(GEOID, year, pct) |>
  pivot_wider(names_from = year, values_from = pct, names_prefix = "pct_") |>
  mutate(
    diff_20yr = round(pct_2020 - pct_2000, 1),
    diff_10yr = round(pct_2020 - pct_2010, 1)
  )

# Join geometries and reproject to WGS84 for Leaflet
fl_map <- left_join(fl_geo, change_hisp, by = "GEOID") |>
  st_transform(4326)
```

## Two Decades of Demographic Change Across Florida

Between 2000 and 2020, Florida's Hispanic population grew rapidly, expanding not just in traditional metro areas like Miami-Dade, but across central Florida (such as Osceola and Orange counties) and southwest Florida.

Hover over any county to see its 20-year path (2000 → 2010 → 2020). Use the layer controls in the top right to switch between the 20-year percentage point increase and the 2020 baseline.

```{r}
#| label: interactive-map
#| message: false
#| echo: false

pal_change <- colorNumeric("YlOrRd", domain = fl_map$diff_20yr)
pal_2020   <- colorNumeric("Blues", domain = fl_map$pct_2020)

leaflet(fl_map, width = "100%", height = "520px") |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    group = "20-Year Change (2000–2020)",
    fillColor = ~pal_change(diff_20yr),
    fillOpacity = 0.7,
    weight = 1,
    color = "white",
    label = ~paste0(NAME, ": +", diff_20yr, " pts (", pct_2000, "% in 2000 → ", pct_2010, "% in 2010 → ", pct_2020, "% in 2020)")
  ) |>
  addPolygons(
    group = "2020 Hispanic %",
    fillColor = ~pal_2020(pct_2020),
    fillOpacity = 0.7,
    weight = 1,
    color = "white",
    label = ~paste0(NAME, ": ", pct_2020, "% Hispanic in 2020")
  ) |>
  addLayersControl(
    baseGroups = c("20-Year Change (2000–2020)", "2020 Hispanic %"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    group = "20-Year Change (2000–2020)",
    pal = pal_change,
    values = ~diff_20yr,
    title = "2000–2020 Change (pts)"
  )
```
